Leader functions and qualities across 59 cultures: an exploratory analysis

### 2019-12-02 leadership_across_cultures_contexts.Rmd

The type of group to which a leader text referred was coded into seven researcher-coded categories: residential groups, kin groups, economic groups, political groups at the community level, political group beyond the community level, military groups, and religious groups. Figure \@ref(fig:groupsex) indicates the proportion of text records that were categorized into each of these groups, broken out by mode of subsistence. We rely on a modified classification of subsistence from the HRAF, collapsing “hunter-gatherers” and “primarily hunter-gatherers” into a single category, “hunter-gatherers,” and by collapsing “agro-pastoralists” and “pastoralists” into a single “pastoralists” category. We renamed the HRAF’s “Other” subsistence category, which is applied to societies with various combinations of hunting, fishing, gathering, pastoralism, and horticulture or intensive agriculture, as “mixed.” This produced a subsistence classification with five groups: Hunter-gatherers, Pastoralists, Horticulturalists, Agriculturalists, and Mixed.

```{r groupsex, fig.width=20, fig.height=10, fig.cap="The distribution of text records by subsistence and group type. Areas of the bars are proportional to the number of text records in that category."}
plot_group_subsis #+ plot_group_sex
```

@@

As Figure \@ref(fig:groupsex) reveals, the distribution of text records by group type varies across subsistence type. Across all subsistence types, the most frequent group type is political groups at the community level, with the exception of agriculturalitsts in which supracommunity political groups are the most frequent group type. Leadership beyond the community level is relatively rare outside of agricultural populations and is especially rare among hunter-gatherers. Leadership within residential groups is most frequently associated with hunter-gatherers (`r hg_residential_pct`% of the total sample of hunter-gatherer text records) and leadership within kin groups is most frequently associated with hunter-gatherers (`r hg_kin_pct`% of the total hunter-gatherers sample) and horticulturalists (`r hort_kin_pct`% of the total horticulturalists sample).

@@

For both leaders and followers, individuals stand to lose what they are able to gain. The clear differences lie in leaders' opportunities for increased social status and matting opportunities. See Figure \@ref(fig:varsupportcb).

@@

<!-- To explore potential population variation in the *Prestige vs. Dominance* and *In-group favoritism vs. Social Status* leader qualities components and the *Social functions* and *Mediation vs. Organization* leader functions components, we fit four hierarchical linear models at the text-record level, using nested random effects of document within culture with each of these components as outcome measures. For predictor variables we relied on our researcher-coded measure of group structure and our modified version of the HRAF subsistence classification (see Figure \@ref(fig:groupsex)).  -->

@@

## The universality of human leadership

Across a broad range of cultures and social contexts leaders are nearly universally high status males described as knowledgeable or intelligent who are experienced and have a demonstrated history of previous accomplishments or special skills. They most frequently resolve conflicts within their respective groups while also providing important social functions, organizing cooperation among followers, and providing counsel or input in followers' affairs. See Figure \@ref(fig:varsupport).

At all subsistence levels leaders emerge across a broad range of group types, including at lower-order social units such as within residential groups and kin groups; within specialized groups such as religious, economic, or military groups; and in political contexts at the community level and beyond. Political leadership at the community level is widespread across all subsistence categories. Economic, military, and religious group leadership, although relatively less frequent, are fairly consistently represented across subsistence levels. See Figure \@ref(fig:groupsex).

## The cultural and context dependency of human leadership

The distribution of text records by group type and subsistence suggests leaders emerge within similar social domains among populations within varied subsistence strategies, but at variable frequencies. Hunter-gatherers and pastoralists, populations more likely to experience high residential mobility, exhibited greater relative frequencies of leadership within residential groups and kin groups. Pastoralists, however, rely in part on a highly defensible resource base (i.e., livestock) and exhibit leadership beyond the community more so than hunter-gatherers and on par with horticulturalists. Pastoral and horticultural populations often experience elevated rates of inter-group conflict relative to mobile hunter-gatherers and these results are consistent with a relationship between increased inter-group conflict and more elaborated political leadership systems.

Figure \@ref(fig:groupsex) is consistent with a negative relationships between political leadership at the community leadership or above and leadership in kin groups or residential groups. One possible interpretation is that as political structures become more institutionalized and economic systems transcend the household, leadership within households and residential groups erodes. It is interesting, however, that even among the agricultural populations that are characterized by leadership systems beyond the community level, leadership within residential and family group persists.

<!-- MAY WANT TO RETHINK INTERPRETATIONS, LOTS OF STORIES HERE -->

## The features of human leadership

@@

>>>>

The type of group to which a leader text referred was coded into seven researcher-coded categories: residential groups, kin groups, economic groups, political groups at the community level, political group beyond the community level, military groups, and religious groups. Figure \@ref(fig:grouptype) indicates the proportion of text records that were categorized into each of these groups, broken out by mode of subsistence. We rely on a modified classification of subsistence from the HRAF, collapsing “hunter-gatherers” and “primarily hunter-gatherers” into a single category, “hunter-gatherers,” and by collapsing “agro-pastoralists” and “pastoralists” into a single “pastoralists” category. We renamed the HRAF’s “Other” subsistence category, which is applied to societies with various combinations of hunting, fishing, gathering, pastoralism, and horticulture or intensive agriculture, as “mixed.” This produced a subsistence classification with five groups: Hunter-gatherers, Pastoralists, Horticulturalists, Agriculturalists, and Mixed.

```{r grouptype, fig.width=20, fig.height=10, fig.cap="The distribution of text records by subsistence and group type. Areas of the bars are proportional to the number of text records in that category."}
plot_group_subsis #+ plot_group_sex
```

@@

As Figure \@ref(fig:grouptype) reveals, the distribution of text records by group type varies across subsistence type. Across all subsistence types, the most frequent group type is political groups at the community level, with the exception of agriculturalitsts in which supracommunity political groups are the most frequent group type. Leadership beyond the community level is relatively rare outside of agricultural populations and is especially rare among hunter-gatherers. Leadership within residential groups is most frequently associated with hunter-gatherers (`r hg_residential_pct`% of the total sample of hunter-gatherer text records) and leadership within kin groups is most frequently associated with hunter-gatherers (`r hg_kin_pct`% of the total hunter-gatherers sample) and horticulturalists (`r hort_kin_pct`% of the total horticulturalists sample).

@@

For both leaders and followers, individuals stand to lose what they are able to gain. The clear differences lie in leaders' opportunities for increased social status and mating opportunities. See Figure \@ref(fig:varsupportcb).

@@

<!-- To explore potential population variation in the *Prestige vs. Dominance* and *In-group favoritism vs. Social Status* leader qualities components and the *Social functions* and *Mediation vs. Organization* leader functions components, we fit four hierarchical linear models at the text-record level, using nested random effects of document within culture with each of these components as outcome measures. For predictor variables we relied on our researcher-coded measure of group structure and our modified version of the HRAF subsistence classification (see Figure \@ref(fig:grouptype)).  -->

@@

## The group context of human leadership

The distribution of text records by group type and subsistence suggests leaders emerge within similar social domains among populations within varied subsistence strategies, but at variable frequencies. See Figure \@ref(fig:grouptype).

### Universal dimensions

In groups of all subsistence types (e.g., hunter-gatherers, pastoralists) leaders emerge across a similar range of group types, including at lower-level social units such as within residential groups and kin groups; within specialized groups such as religious, economic, or military groups; and in political contexts at the community and supracommunity levels. A large percentage of text records in all subsistence groups document political leadership at the community level. Leadership at the supracommunity level is also common across all subsistence groups save hunter-gatherers, which are largely autonomous. Economic, military, and religious group leadership, although relatively less frequent, are fairly consistently represented across subsistence levels.

### Variable dimensions

Leadership in residential subgroups (e.g., hunting party, work group) was relatively common among hunter-gatherers but relatively rare among agriculturists. Leadership at the surpacommunity level, in contrast, was relatively rare among hunter-gatherers but quite common among agriculturists.

Hunter-gatherers and pastoralists, populations more likely to experience high residential mobility, exhibited greater relative frequencies of leadership within residential groups and kin groups. Pastoralists, however, rely in part on a highly defensible resource base (i.e., livestock) and exhibit leadership beyond the community more so than hunter-gatherers and on par with horticulturalists. Pastoral and horticultural populations often experience elevated rates of inter-group conflict relative to mobile hunter-gatherers and these results are consistent with a relationship between increased inter-group conflict and more elaborated political leadership systems.

Figure \@ref(fig:grouptype) is consistent with a negative relationships between political leadership at the community leadership or above and leadership in kin groups or residential groups. One possible interpretation is that as political structures become more institutionalized and economic systems transcend the household, leadership within households and residential groups erodes. It is interesting, however, that even among the agricultural populations that are characterized by leadership systems beyond the community level, leadership within residential and family group persists.

## Leader functions and qualities

Our analysis identified several leader functions and qualities that were near cultural universals, and others that were relatively rare. See Figure \@ref(fig:varsupport).

### Universal dimensions

Across a broad range of cultures and social contexts leaders are nearly universally high status males who are described as knowledgeable or intelligent and who are experienced and have a demonstrated history of previous accomplishments or special skills. They most frequently resolve conflicts within their respective groups while also providing important social functions, organizing cooperation among followers, providing counsel or input in followers' affairs, and distributing resources. See Figure \@ref(fig:varsupport).

### Variable dimensions

Several functions and qualities that are theoretically important were relatively rare in the ethnographic record. In the literature on leadership among non-human animals, for example, leaders commonly initiate group movement [@couzin_effective_2005], and "movement" was one of four leadership dimensions that @smith_leadership_2016 identified as comparable among humans and non-human animals, yet movement or migration was identified as a leader function in only `r movement_pct`% of cultures. Several theories of leadership also emphasize the moral authority of leaders [e.g., @xx; @yy], yet leaders were seen as sources of moral authority or "fair" in only a minority of cultures (`r moral_pct`% and `r fairness_pct`% of cultures, respectively). The "Dominance" theory of leadership [@xx; @yy] emphasizes leaders' ability to instill fear, but in only `r feared_pct`% of cultures were leaders characterized as feared. Finally, a large literature on leadership in industrialized societies highlights the attractiveness of leaders, but leaders were characterized as attractive in only `r attractive_pct`% of cultures.

## Clusters of leader functions and qualities

@@

## Synthesis
